Temporal stability in human interaction networks: sector sizes, topological prominence and time activity

### 2015-10-04 paper.tex
feita revisao quase ateh finalizar seccionamento

@@

%This is a way to enhance the reliability of the methods,
%of algorithmic routines, of data consistency,
%and of the results themselves.
%Also, we believe that this practice raises the scientific contribution of the work,
%handing not only the framework and the results, 
%but the exact data and processes that render them~\cite{openSci}.
\section{Results and discussion}\label{sec:results}
%Remarkable features from the analysis of the four email lists are:
%\begin{itemize}
%    \item The activity along time is practically the same for all lists, thus suggesting stable patterns.
%    \item The fraction of participants in each Erd\"os sector is stable along time and can be determined even with very few messages 
%    \item The topological metrics combine into principal components in PCA in the same way for all lists and all snapshots (). 
%        \item Symmetry measures of the topology, as defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
%    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
%\end{itemize}
\subsection{Activity along time}\label{constDisc}
The observed activity along time, in terms of seconds, minutes, hours, days and months,  is practically the same for all lists.
Histograms in each time scale were computed as were circular average values and their dispersion.
We chose to provide detailed values in Table~\ref{tab:circ}-\ref{tab:min2} because these numbers can actually be used for characterizing nodes (participants) in other networks, and networks themselves, as they are independent of the network under analysis. For example, they may serve for identification of outliers in a community.

\begin{table}
	\caption{The rescaled circular mean $\theta_\mu'$ and the circular dispersion $\delta(z)$ described in Section~\ref{sec:mtime}.
		This typical table was constructed using all LAD list messages, and the results are the same for other lists, as shown in Section~\ref{si:circ} of the Supporting Information.
	The most uniform distribution of activity was found in seconds and minutes, where the mean has little meaning.
	Hours of the day exhibited the most concentrated activity (lowest $\delta(z)$), with mean between 14h and 15h ($\theta'=-9.61$).
	Weekdays, month days and months have mean near zero (i.e. near the beginning of the week, month and year) and high dispersion.}
	\begin{center}
		\begin{tabular}{ |l|| c|c| }
			\hline
			%scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
			%& $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$  \\ \hline\hline
			{\bf scale} & {\bf mean} \boldsymbol{$\theta_\mu'$} & {\bf dispersion} \boldsymbol{$\delta(z)$ }  \\ \hline\hline
			\input{tables/tab2TimeLAD___}
		\end{tabular}
	\end{center}
	\label{tab:circ}
\end{table}

\begin{table}
	\caption{Activity percentages along the hours of the day for the CPP list. Nearly identical distributions are found on other lists as shown in Section~\ref{si:hours} of the Supporting Information. Higher activity was observed between noon and 6pm, followed by the time period between 6pm and midnight. Around 2/3 of the whole activity takes place from noon to midnight. Nevertheless, the activity peak occurs around midday, with a slight skew toward one hour before noon.}
	\footnotesize
	\input{tables/tabHoursCPP_}
	\label{tab:hin}
\end{table}

\begin{table}
	\caption{Activity percentages along the days of the week for the four email lists. Higher activity was observed during weekdays, with a decrease of activity on weekends of at least one third and two thirds in extreme cases.}
	\begin{center}
		\begin{tabular}{ | l |  c | c | c | c | c |   c | c |}
			\hline
			& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
			\input{tables/tabWeekdays}
		\end{tabular}
	\end{center}
	\label{tab:win}
\end{table}
In the scale of seconds and minutes, activity obeys a homogeneous pattern, with the messages being slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}. In the networks, $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$ while simulations reach these values but have on average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected. Circular dispersion is maximized and the mean has little meaning as indicated in Table~\ref{tab:circ}. As for the hours of the day, an abrupt peak appeared around 11am with the most active period being the afternoon. Days of the week revealed a decrease between one third and two thirds of activity on weekends. Days of the month were regarded as homogeneous with an inconclusive slight tendency of the first week being more active. Months of the year revealed patterns matching usual work and academic calendars. The time period examined here was not sufficient for the analysis of activity along the years. These patterns are exemplified in Tables~\ref{tab:hin}-\ref{tab:min2}.

@@

\begin{table}
	\caption{Activity percentages of the months along the year from LAU list. Activity is usually concentrated in Jun-Aug and/or in Dec-Mar (see Section~\ref{si:months} of the Supporting Information). These observations fit academic calendars, vacations and end-of-year holidays.}
	\footnotesize
	\input{tables/tabMonthsLAU}
	\label{tab:min2}
\end{table}
\subsection{Stability of principal components and the prevalence of symmetry over clusterization for dispersion}\label{prevalence}
The topology was analyzed using standard, well-established metrics of centrality and clustering.
We also introduced symmetry metrics given the evidence of their importance in social contexts~\cite{newmanEvolving}.
The contribution of each metric to the variance is very similar for all the networks, and did not vary with time.
In applying PCA to the snapshots, the contribution of each metric to the principal components presents very small standard deviation. Table~\ref{tab:pcain} exemplifies the principal components formation with all the metrics considered for the MET email list. Similar results are presented in Section~\ref{si:pcat} of the Supporting Information for the other lists, with separate consideration of strategic combinations of metrics.

@@

The first principal component is an average of centrality metrics: degrees, strengths and betweenness centrality. Therefore, all of these centrality measurements are equally important for characterizing the networks. On one hand, the relevance of all centrality metrics is not surprising since they may be highly correlated. The degree and strength, for instance, are highly correlated, with Spearman correlation coefficient $\in [0.95,1]$ and Pearson coefficient $\in [0.85,1)$ for $ws>1000$. On the other hand, each of these metrics is related to a different participation characteristic, and their equal relevance is noticeable.
The clustering coefficient is presented in almost perfect orthogonality to centrality metrics.
\begin{figure*} 
	\centering
	\includegraphics[width=.6\textwidth,height=10cm]{figs/im13PCAPLOT_}
	\caption{The first plot shows degree versus clustering coefficient. This typical pattern is well known, since high clustering is more incident in vertices with lower degrees. The second plot is analogous but the first component is an average of centrality metrics. The second component remains related to the clustering coefficient. The third plot exhibits the greater dispersion in the symmetry-related second component. In this case, the clustering coefficient is only relevant for the third component. This greater dispersion suggests that symmetry-related metrics are more powerful for characterizing interaction networks than the clustering coefficient, especially for hubs and intermediary vertices.
	This figure was obtained with a snapshot of the LAU list in a window size of $ws = 1000$ messages. Similar structures were observed in all window sizes $ws\;\in\;[500,10000]$ and for networks of other email lists, which points to a common relationship between the metrics of degrees, strengths and betweenness centrality, the symmetry-related metrics and clustering coefficient.}
	\label{fig:sym}
\end{figure*}

Dispersion was more prevalent in symmetry-related metrics than for the clustering coefficient, as indicated in Table~\ref{tab:pcain}. This is also illustrated in Figure~\ref{fig:sym}, where each vertex is colored according to the sector they belong to. As expected, peripheral vertices have very low values in the first component (centrality related) and greater dispersion in the third component (clustering related).
The PCA plot in the third system of Figure~\ref{fig:sym}, where all metrics are considered, reflects the relevance of the symmetry-related metrics for the variance.
We conclude that the latter metrics can be more meaningful in characterizing interaction networks (and their participants) than the clustering coefficient, especially for hubs and intermediary vertices.

The relative importance of the topological metrics was also observed for the additional 12 networks from Facebook, Twitter and Participabr. With the exception of two of these networks, the overall behavior was maintained in that centrality measurements were found to be the most relevant to explain network topology, followed by symmetry-related metrics and then clustering coefficient. The results are given in Tables~\ref{tab:pcaE1F},~\ref{tab:pcaE1I},~\ref{tab:pcaE2},~\ref{tab:pcaE3} of the Supporting Information. There are larger differences between two of these networks than between two (GMANE) email networks, as the latter were much more regular.

@@

The distribution of vertices in the hubs, intermediary, periphery Erd\"os sectors is remarkably stable along time, provided that a sufficiently large sample of 200 or more messages is considered. 
Moreover, the same distribution applies to the networks of all email lists analyzed, as demonstrated in Figure~\ref{fig:sectIL} and in Section~\ref{si:frac} of the Supporting Information. 
Activity is highly concentrated on the hubs, while a very large number of peripheral vertices contribute to only a fraction of the activity. This is expected for a system with a scale-free profile, as confirmed by the data in Table~\ref{autores} for the distribution of activity among participants.

Typically, $\approx [3-12]\%$ of the vertices are found to be hubs, $\approx [15-45]\%$ are intermediary and $\approx [44-81]\%$ are peripheral, which is consistent with the literature~\cite{secFree}.
These results hold for the total, in and out degrees and strengths.
Stable distributions can also be obtained for 100 or less messages if classification of the three sectors is performed with one of the compound criteria established in Section~\ref{sectioning}. 
The networks hold their basic structure with as few as 10-50 messages; concentration of activity and the abundance of low-activity participants take place even with very few messages, which is highlighted in Section~\ref{si:frac} of the Supporting Information.
A minimum window size for the observation of more general properties might be inferred by monitoring the giant component and the degeneration of the Erd\"os sectors.

In order to verify the possible generality of these findings, we obtained the Erd\"os sectors of 12 networks from Facebook, Twitter and Participabr. The results are given in Table~\ref{tab:secE} in the Supporting Information, which indicate that the percentages of hubs, intermediary and peripheral nodes are essentially the same as for the email lists.

@@

\begin{table}[h]
	\caption{Distribution of activity among participants. The first column presents the percentage of messages sent by the most active participant. The column for the first quartile ($1Q$) shows the minimum percentage of participants responsible for at least 25\% of total messages. Similarly, the column for the first three quartiles $1-3Q$ gives the minimum percentage of participants responsible for 75\% of total messages. The last decile $-10D$ column brings the maximum percentage of participants responsible for 10\% of messages.}
	\begin{center}
		\begin{tabular}{ | l ||  c | c | c | c | }
			\hline
			list & hub & $ 1Q $ & $ 1-3Q $ & $-10D$ \\ \hline
			\input{tables/userTab}
		\end{tabular}
	\end{center}
	\label{autores}
\end{table}

>>>>

These measures are used both for the Erd\"os sectioning (described in Secion~\ref{sectioning}) and for performing PCA (as described in Section~\ref{sec:pca}).

@@

%This is a way to enhance the reliability of the methods,
%of algorithmic routines, of data consistency,
%and of the results themselves.
%Also, we believe that this practice raises the scientific contribution of the work,
%handing not only the framework and the results, 
%but the exact data and processes that render them~\cite{openSci}.
\section{Results and discussion}\label{sec:results}
%Remarkable features from the analysis of the four email lists are:
%\begin{itemize}
%    \item The activity along time is practically the same for all lists, thus suggesting stable patterns.
%    \item The fraction of participants in each Erd\"os sector is stable along time and can be determined even with very few messages 
%    \item The topological metrics combine into principal components in PCA in the same way for all lists and all snapshots (). 
%        \item Symmetry measures of the topology, as defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
%    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
%\end{itemize}
\subsection{Activity along time}\label{constDisc}
The observed activity along time, in terms of seconds, minutes, hours, days and months,  is practically the same for all lists.
Histograms in each time scale were computed as were circular average and dispersion values.
We provide detailed values in Table~\ref{tab:circ}-\ref{tab:min2} because they can
be used for characterizing nodes (participants) in these and other networks,
and networks themselves.
For example, they may serve for identification of outliers in a community.

\begin{table}
	\caption{The rescaled circular mean $\theta_\mu'$ and the circular dispersion $\delta(z)$ described in Section~\ref{sec:mtime}.
		This typical table was constructed using all LAD list messages, and the results are the same for other lists, as shown in Section~\ref{si:circ} of the Supporting Information document.
	The most uniform distribution of activity was found in seconds and minutes, where the mean has little meaning.
	Hours of the day exhibited the most concentrated activity (lowest $\delta(z)$), with mean between 14h and 15h ($\theta'=-9.61$).
	Weekdays, month days and months have mean near zero (i.e. near the beginning of the week, month and year) and high dispersion.}
	\begin{center}
		\begin{tabular}{ |l|| c|c| }
			\hline
			%scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
			%& $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$  \\ \hline\hline
			{\bf scale} & {\bf mean} \boldsymbol{$\theta_\mu'$} & {\bf dispersion} \boldsymbol{$\delta(z)$ }  \\ \hline\hline
			\input{tables/tab2TimeLAD___}
		\end{tabular}
	\end{center}
	\label{tab:circ}
\end{table}

\begin{table}
	\caption{Activity percentages along the hours of the day for the CPP list. Nearly identical distributions are found on other lists as shown in Section~\ref{si:hours} of the Supporting Information document. Higher activity was observed between noon and 6pm, followed by the time period between 6pm and midnight. Around 2/3 of the whole activity takes place from noon to midnight. Nevertheless, the activity peak occurs around midday, with a slight skew toward one hour before noon.}
	\footnotesize
	\input{tables/tabHoursCPP_}
	\label{tab:hin}
\end{table}

\begin{table}
	\caption{Activity percentages along the days of the week for the four email lists. Higher activity was observed during weekdays, with a decrease of activity on weekends of at least one third and two thirds in extreme cases.}
	\begin{center}
		\begin{tabular}{ | l |  c | c | c | c | c |   c | c |}
			\hline
			& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
			\input{tables/tabWeekdays}
		\end{tabular}
	\end{center}
	\label{tab:win}
\end{table}
In the scale of seconds and minutes, activity obeys a homogeneous pattern, with the messages being slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}. In the networks, $\frac{min(incidence)}{max(incidence)} \in (0.784,.794)$ while simulations reach these values but have on average more discrepant higher and lower peaks $\xi=\frac{min(incidence')}{max(incidence')} \Rightarrow \mu_\xi=0.7741 \text{ and } \sigma_\xi=0.02619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform.
In these cases, the circular dispersion is maximized and the mean has little meaning as indicated in Table~\ref{tab:circ}. As for the hours of the day, an abrupt peak appeared around 11am with the most active period being the afternoon. Days of the week revealed a decrease between one third and two thirds of activity on weekends. Days of the month were regarded as homogeneous with an inconclusive slight tendency of the first week being more active. Months of the year revealed patterns matching usual work and academic calendars. The time period examined here was not sufficient for the analysis of activity along the years. These patterns are exemplified in Tables~\ref{tab:hin}-\ref{tab:min2}.

@@

\begin{table}
	\caption{Activity percentages of the months along the year from LAU list. Activity is usually concentrated in Jun-Aug and/or in Dec-Mar (see Section~\ref{si:months} of the Supporting Information). These observations fit academic calendars, vacations and end-of-year holidays.}
	\footnotesize
	\input{tables/tabMonthsLAU}
	\label{tab:min2}
\end{table}
\subsection{Stability of principal components and the prevalence of symmetry over clusterization for dispersion}\label{prevalence}
%The topology was analyzed using standard, well-established metrics of centrality and clustering.
%We also introduced symmetry metrics given the evidence of their importance in social contexts~\cite{newmanEvolving}.
The contribution of each metric to the variance is very similar for all the networks and along time.
In applying PCA to the snapshots, the contribution of each metric to the principal components presents very small standard deviation.
Table~\ref{tab:pcain} exemplifies the formation of principal components for the MET email list.
Similar results are presented in Sections~\ref{si:pcat} and~\ref{si:ext}
of the Supporting Information document for the other email lists (benchmarks) and other interaction networks,
with the consideration of strategic combinations of metrics.

@@

The first principal component is an average of centrality metrics: degrees, strengths and betweenness centrality. Therefore, all of these centrality measurements are equally important for characterizing the networks. On one hand, the relevance of all centrality metrics is not surprising since they are highly correlated, e.g.the degree and strength have Spearman correlation coefficient $\in [0.95,1]$ and Pearson coefficient $\in [0.85,1)$ for $ws>1000$.
	On the other hand, each of these metrics is related to a different participation characteristic, and their equal relevance is noticeable.
The clustering coefficient is presented in almost perfect orthogonality to centrality metrics.
\begin{figure*} 
	\centering
	\includegraphics[width=.6\textwidth,height=10cm]{figs/im13PCAPLOT_}
	\caption{The first plot shows degree versus clustering coefficient. This typical pattern is well known with high clustering more incident in vertices with lower degrees. 
		The second plot is analogous but the abscissas is an average of centrality metrics.
		The third plot exhibits the greater dispersion of the symmetry-related ordinates.
		This greater dispersion suggests that symmetry-related metrics are more powerful, with respect to dispersion,
		for characterizing interaction networks than the clustering coefficient,
		especially for hubs and intermediary vertices.
		This figure was obtained with a snapshot of the LAU list in a window size of $ws = 1000$ messages.
		Similar structures were observed in all window sizes $ws\;\in\;[500,10000]$, in networks derived from email lists,
		and in networks from Facebook, Twitter and Participabr,
		which suggests a common relationship between the metrics of degrees, strengths and betweenness centrality,
		the symmetry-related metrics and clustering coefficient.}
	\label{fig:sym}
\end{figure*}

Dispersion was more prevalent in symmetry-related metrics than for the clustering coefficient, as indicated in Table~\ref{tab:pcain}. This is also illustrated in Figure~\ref{fig:sym}, where each vertex is colored according to the sector they belong to. As expected, peripheral vertices have very low values in the first component (centrality related) and greater dispersion in the third component (clustering related).
The PCA plot in the third system of Figure~\ref{fig:sym}, where all metrics are considered, reflects the relevance of the symmetry-related metrics for the variance.
We conclude that the symmetry metrics is more useful, with respect to the dispersion in the topological space,
in characterizing interaction networks (and their participants) than the clustering coefficient,
especially for hubs and intermediary vertices.

The relative importance of the topological metrics was also observed for the additional 12 networks from Facebook, Twitter and Participabr.
With the exception of two of these networks, the overall behavior was maintained in that centrality measurements 
were found prevalent in the first principal component,
followed by symmetry-related metrics on the second principal component and then clustering coefficient on the third principal component.
The results are given in Tables~\ref{tab:pcaE1F},~\ref{tab:pcaE1I},~\ref{tab:pcaE2},~\ref{tab:pcaE3} of the Supporting Information document.
Larger variability was found among such networks, which motivated the
use of interaction networks derived from email lists for benchmarks.

@@

The distribution of vertices in the hubs, intermediary, periphery Erd\"os sectors is remarkably stable along time, provided that a sufficiently large sample of 200 or more messages is considered. 
Moreover, the same distribution applies to the networks of all email lists analyzed,
as demonstrated in Figure~\ref{fig:sectIL} and in Section~\ref{si:frac} of the Supporting Information document. 
Activity is highly concentrated on the hubs, while a very large number of peripheral vertices contribute to only a fraction of the activity. This is expected for a system with a scale-free profile, as confirmed by the data in Table~\ref{autores} for the distribution of activity among participants.

Typically, $\approx [3-12]\%$ of the vertices are hubs, $\approx [15-45]\%$ are intermediary and $\approx [44-81]\%$ are peripheral, which is consistent with the literature~\cite{secFree}.
These results hold for the total, in and out degrees and strengths.
Stable distributions can also be obtained for 100 or less messages if classification of the three sectors is performed with one of the compound criteria established in Section~\ref{sectioning}. 
The networks hold their basic structure with as few as 10-50 messages,
i.e. concentration of activity and the abundance of low-activity participants take place even with very few messages,
which is highlighted in Section~\ref{si:frac} of the Supporting Information document.
A minimum window size for the observation of more general properties might be inferred by monitoring 
both the giant component and the degeneration of the Erd\"os sectors.

In order to support hypotheses about the generality of these findings,
we obtained the Erd\"os sectors of 12 networks from Facebook, Twitter and Participabr.
The results are given in Table~\ref{tab:secE} of the Supporting Information,
which indicate that the percentages of hubs, intermediary and periphery nodes are essentially the same as for the email lists.

@@

\begin{table}[h]
	\caption{Distribution of activity among participants. The first column presents the percentage of messages sent by the most active participant. The column for the first quartile ($1Q$) shows the minimum percentage of participants responsible for at least 25\% of total messages. Similarly, the column for the first three quartiles $1-3Q$ gives the minimum percentage of participants responsible for 75\% of total messages. The last decile $-10D$ column exposes the maximum percentage of participants responsible for 10\% of messages.}
	\begin{center}
		\begin{tabular}{ | l ||  c | c | c | c | }
			\hline
			list & hub & $ 1Q $ & $ 1-3Q $ & $-10D$ \\ \hline
			\input{tables/userTab}
		\end{tabular}
	\end{center}
	\label{autores}
\end{table}
